---
title: "Debt Investment Analysis"
author: "student"
date: "2025-04-13"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r }
# Load necessary libraries
library(tidyverse)
library(lubridate)
library(kableExtra)
library(ggplot2)
library(scales)
library(dplyr)

```


```{r}
# Load data
accounts_data <- read_csv(choose.files())
payments <- read_csv(choose.files())

```


```{r }
# Merge datasets
combined_data <- accounts_data %>%
  left_join(payments, by = "uuid")
head(combined_data)

# Data cleaning and preparation
combined_data1 <- combined_data %>%
  mutate(
    activation_date = as.Date(activation_date, format = "%m/%d/%Y"),
    date_of_data = as.Date(date_of_data, format = "%m/%d/%Y"),
    repmt_date = as.Date(repmt_date, format = "%m/%d/%Y"),
    financed_amount = total_cost - downpmt_cost,
    upfront_days = downpmt_cost / receivable_per_day,
    contract_weeks = contract_days / 7,
    age_months = as.numeric(date_of_data - activation_date) / 30,
    days_since_last_payment = as.numeric(date_of_data - repmt_date),
    amount_paid = ifelse(is.na(repmt_amt), 0, repmt_amt),
    is_defaulted = ifelse(days_since_last_payment >= 90 | age_months > 36, TRUE, FALSE)

  )
colnames(combined_data1)

```
```{r}
sapply(combined_data1, function(x) sum(is.na(x)))

```


```{r}
# Identify defaulted accounts
combined_data1 %>%
  mutate(
    repmt_date = as.Date(repmt_date, format = "%m/%d/%Y"),
    days_since_last_payment = as.numeric(date_of_data - repmt_date),
    amount_paid = ifelse(is.na(repmt_amt), 0, repmt_amt),
    is_defaulted = ifelse(days_since_last_payment >= 90 | age_months > 36, TRUE, FALSE)
  ) %>%
  mutate(
    is_defaulted = case_when(
      days_since_last_payment >= 90 ~ TRUE,
      age_months > 36 ~ TRUE,
      TRUE ~ FALSE
    )
  )
```


```{r PART A, echo=FALSE}
# metrics by product group
product_analysis <- combined_data1 %>%
  group_by(product_name) %>%
  summarise(
    avg_financed_amount = mean(financed_amount, na.rm = TRUE),
    avg_deposit_amount = mean(downpmt_cost, na.rm = TRUE),
    avg_upfront_days = mean(upfront_days, na.rm = TRUE),
    avg_contract_length_weeks = mean(contract_weeks, na.rm = TRUE),
    count = n(),
    total_sales_value = sum(total_cost, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    prop_by_count = count / sum(count),
    prop_by_value = total_sales_value / sum(total_sales_value)
  ) %>%
  select(
    product_name,
    avg_financed_amount,
    avg_deposit_amount,
    avg_upfront_days,
    avg_contract_length_weeks,
    prop_by_count,
    prop_by_value,
    total_sales_value
  )

# Table Format
product_analysis %>%
  mutate(
    avg_financed_amount = dollar(avg_financed_amount, accuracy = 1),
    avg_deposit_amount = dollar(avg_deposit_amount, accuracy = 1),
    avg_upfront_days = round(avg_upfront_days, 1),
    avg_contract_length_weeks = round(avg_contract_length_weeks, 1),
    prop_by_count = percent(prop_by_count, accuracy = 0.1),
    prop_by_value = percent(prop_by_value, accuracy = 0.1),
    total_sales_value = dollar(total_sales_value, accuracy = 1)
  ) %>%
  kable(
    col.names = c(
      "Product", "Avg Financed Amount", "Avg Deposit Amount", 
      "Avg Upfront Days", "Avg Contract Length (weeks)", 
      "% of Total Sales (count)", "% of Total Sales (value)",
      "Total Sales Value"
    ),
    caption = "Product Group Analysis"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r PART B}
# Ensure date columns are in Date format
combined_data1 <- combined_data1 %>%
  mutate(
    activation_date = as.Date(activation_date),
    date_of_data = as.Date(date_of_data),
    repmt_date = as.Date(repmt_date, format = "%m/%d/%Y")
  )

#  days since activation and days since last payment
portfolio_data <- combined_data1 %>%
  mutate(
    days_since_activation = as.numeric(date_of_data - activation_date),
    age_months = days_since_activation / 30
  ) %>%
  group_by(uuid) %>%
  mutate(
    last_payment_date = max(repmt_date, na.rm = TRUE),
    total_paid = sum(repmt_amt, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    days_since_last_payment = as.numeric(date_of_data - last_payment_date),
    outstanding_balance = pmax(total_cost - total_paid, 0),
    is_default = ifelse(
      days_since_last_payment >= 90 | age_months > 36,
      TRUE,
      FALSE
    )
  )

# portfolio metrics
portfolio_metrics <- list(
  total_portfolio_outstanding = sum(portfolio_data$outstanding_balance, na.rm = TRUE),
  defaulted_balance = sum(portfolio_data$outstanding_balance[portfolio_data$is_default], na.rm = TRUE),
  active_balance = sum(portfolio_data$outstanding_balance[!portfolio_data$is_default], na.rm = TRUE)
) %>%
  as_tibble() %>%
  mutate(
    default_ratio = defaulted_balance / total_portfolio_outstanding
  )

#  formatted table
portfolio_metrics %>%
  mutate(
    total_portfolio_outstanding = dollar(total_portfolio_outstanding, accuracy = 1),
    defaulted_balance = dollar(defaulted_balance, accuracy = 1),
    active_balance = dollar(active_balance, accuracy = 1),
    default_ratio = percent(default_ratio, accuracy = 0.1)
  ) %>%
  kable(
    col.names = c(
      "Total Portfolio Outstanding", 
      "Defaulted Balance", 
      "Active Balance",
      "Default Ratio"
    ),
    caption = "Key Portfolio Metrics"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

```{r PART C_analysis}
# Create activation cohorts
cohort_analysis <- portfolio_data %>%
  mutate(
    activation_quarter = quarter(activation_date, with_year = TRUE),
    activation_cohort = paste0(
      year(activation_date), 
      " Q", 
      quarter(activation_date)
    )
  ) %>%
  group_by(activation_cohort, country) %>%
  summarise(
    avg_outstanding_balance = mean(outstanding_balance, na.rm = TRUE),
    default_rate = mean(is_default, na.rm = TRUE),
    avg_pct_paid = mean(total_paid / total_cost, na.rm = TRUE),
    n_accounts = n(),
    .groups = 'drop'
  )

# Visualize repayment by cohort
ggplot(cohort_analysis, aes(x = activation_cohort, y = avg_pct_paid, fill = country)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Percentage Paid by Activation Cohort",
    x = "Activation Cohort",
    y = "Average Percentage of Contract Paid",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Visualize default rates by cohort
ggplot(cohort_analysis, aes(x = activation_cohort, y = default_rate, fill = country)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Default Rate by Activation Cohort",
    x = "Activation Cohort",
    y = "Default Rate",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r PART C}
# Create activation cohorts by quarter
cohort_analysis <- combined_data1 %>%
  mutate(
    activation_quarter = quarter(activation_date, with_year = TRUE),
    activation_month = floor_date(activation_date, "month")
  )

# Calculate repayment rates by cohort
repayment_rates <- cohort_analysis %>%
  group_by(activation_quarter, product_name) %>%
  summarise(
    total_contracts = n(),
    default_rate = mean(is_defaulted, na.rm = TRUE),
    avg_pct_paid = mean(amount_paid / financed_amount, na.rm = TRUE),
    .groups = "drop"
  )

# Visualization: Default rates by cohort and product
ggplot(repayment_rates, aes(x = activation_quarter, y = default_rate, color = product_name)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Default Rates by Activation Cohort and Product",
    x = "Activation Quarter",
    y = "Default Rate",
    color = "Product"
  ) +
  theme_minimal()
```


## Key Findings and Recommendations:

The analysis reveals several critical insights about the distributed energy company's portfolio. First, product differentiation significantly impacts financial performance, with SunPump representing 68.4% of total sales value despite comprising only 29.1% of contracts by count. This high-value product has substantially different terms (average financed amount of 155,513 with 60 upfront days) compared to other offerings.The portfolio shows a 25.9 % and 3.07 billion in defaulted receivables, indicating notable credit risk. Cohort analysis demonstrates improving repayment trends in recent quarters, suggesting underwriting or operational enhancements, though certain products like SunExtra + Radio show elevated default rates in specific markets.

For the investment team, we recommend: 

(1) implementing product-specific risk pricing, particularly for SunPump given its outsized portfolio contribution and unique risk profile; 

(2) establishing geographic concentration limits for markets showing persistently high default rates; and

(3) structuring the investment with performance triggers tied to cohort-level repayment metrics to mitigate risks from potential vintage quality fluctuations. 

The improving repayment trends in recent cohorts are encouraging, but continued monitoring of product mix evolution and geographic expansion will be essential for risk management. A tiered investment approach that segments exposure by product risk category could optimize risk-adjusted returns.
